Supplementary Materials for CBEA: Competitive Balances for Taxonomic Enrichment Analysis

### 2022-04-30 cbea_revised_v4.tex
V4

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{CBEA: Competitive balances for taxonomic enrichment analysis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Quang P. Nguyen\textsuperscript{1,2},
Anne G. Hoen\textsuperscript{1,2\ddag},
H. Robert Frost\textsuperscript{1*\ddag\dag},
\\
\bigskip
\textbf{1} Department of Biomedical Data Science, Geisel School of Medicine at Dartmouth College, Hanover, New Hampshire, United States
\\
\textbf{2} Department of Epidemiology, Geisel School of Medicine at Dartmouth College, Hanover, New Hampshire, United States
\\
\bigskip

@@

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
\ddag These authors jointly supervised this work.\\
\dag Corresponding author
% Current address notes
%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

@@

\section*{Results}
In this section, we present results for evaluating statistical significance, phenotype relevance, and predictive performance. In addition to real data, we also evaluated models based on parametric simulations. Results can be found in the Supplemental Materials (\nameref{S1_Fig}, \nameref{S2_Fig}, \nameref{S3_Fig}, \nameref{S4_Fig}, and \nameref{S5_Fig}).

@@

For single-sample analyses, we demonstrated that the CBEA approach (unadjusted with mixture normal parametric assumption) was able to control for type I error at the nominal level of 0.05 under the global null (Fig~\ref{fig:2}), while also demonstrating adequate power (Fig~\ref{fig:4}). This performance is consistent across different set sizes as well as our prior distributional fit analyses (Fig~\ref{fig:1}), in which the mixture normal displayed superior fit to the null distribution. Unfortunately, other variants of CBEA demonstrated neither good type I error control nor power. Interestingly, while the adjusted methods showed poor performance in real data evaluations (Fig~\ref{fig:2}), in simulation studies (\nameref{S1_Fig}, \nameref{S3_Fig}) these approaches were able to control for type I error well with the trade-off of much lower power. 
For the population-level inference task, CBEA also performed very well. Under the permutation global null, representing genera abundance using CBEA scores in combination with Welch's t-test controls for type I error at the correct $\alpha$ threshold while also keeping respectable power. Since the population level enrichment test is equivalent to a differential abundance test using set-based features, we compared the CBEA approach against using element-wise summations with corncob \cite{martin2020} and DESeq2 \cite{love2014} to test for set-level differential abundance. We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome taxonomic data \cite{mcmurdie2014}. Alternately, corncob is a newer method developed specifically for microbiome taxonomic data sets, which models taxonomic counts directly using a beta-binomial distribution instead of relying on normalization via size factor estimation. We observed that using this approach resulted in an inflated type I error compared to all variants of CBEA (Fig~\ref{fig:2}), yet did not improve power (Fig~\ref{fig:4}). Results for CBEA approaches were replicated in simulation analyses, however, for corncob and DESeq2 we observed an opposite effect: in simulation experiments, both methods show good type I error control but low power (\nameref{S2_Fig}, \nameref{S4_Fig}).

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{CBEA: Competitive balances for taxonomic enrichment analysis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Quang P. Nguyen\textsuperscript{1,2},
Anne G. Hoen\textsuperscript{1,2\dag},
H. Robert Frost\textsuperscript{1*\dag},
\\
\bigskip
\textbf{1} Department of Biomedical Data Science, Geisel School of Medicine at Dartmouth College, Hanover, New Hampshire, United States
\\
\textbf{2} Department of Epidemiology, Geisel School of Medicine at Dartmouth College, Hanover, New Hampshire, United States
\\
\bigskip

@@

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
\dag These authors jointly supervised this work.\\
% Current address notes
%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

@@

\section*{Results}
In this section, we present results for evaluating statistical significance, phenotype relevance, and predictive performance. In addition to real data, we also evaluated models based on parametric simulations. Results can be found in the Supplemental Materials (S1-S5 Figs).

@@

For single-sample analyses, we demonstrated that the CBEA approach (unadjusted with mixture normal parametric assumption) was able to control for type I error at the nominal level of 0.05 under the global null (Fig~\ref{fig:2}), while also demonstrating adequate power (Fig~\ref{fig:4}). This performance is consistent across different set sizes as well as our prior distributional fit analyses (Fig~\ref{fig:1}), in which the mixture normal displayed superior fit to the null distribution. Unfortunately, other variants of CBEA demonstrated neither good type I error control nor power. Interestingly, while the adjusted methods showed poor performance in real data evaluations (Fig~\ref{fig:2}), in simulation studies (S1 and S3 Figs) these approaches were able to control for type I error well with the trade-off of much lower power. 
For the population-level inference task, CBEA also performed very well. Under the permutation global null, representing genera abundance using CBEA scores in combination with Welch's t-test controls for type I error at the correct $\alpha$ threshold while also keeping respectable power. Since the population level enrichment test is equivalent to a differential abundance test using set-based features, we compared the CBEA approach against using element-wise summations with corncob \cite{martin2020} and DESeq2 \cite{love2014} to test for set-level differential abundance. We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome taxonomic data \cite{mcmurdie2014}. Alternately, corncob is a newer method developed specifically for microbiome taxonomic data sets, which models taxonomic counts directly using a beta-binomial distribution instead of relying on normalization via size factor estimation. We observed that using this approach resulted in an inflated type I error compared to all variants of CBEA (Fig~\ref{fig:2}), yet did not improve power (Fig~\ref{fig:4}). Results for CBEA approaches were replicated in simulation analyses, however, for corncob and DESeq2 we observed an opposite effect: in simulation experiments, both methods show good type I error control but low power (S2 and S4 Figs).
